
%\def\filepath{C:/Users/holden-lee/Dropbox/Math/templates}

\input{templates/packages_book.tex}
\input{templates/theorems_with_boxes.tex}
\input{templates/macros.tex}
\input{templates/formatting.tex}
\input{templates/other.tex}
\input{templates/theorem_num.tex}

\def\name{Mathematical Physics}


\pagestyle{fancy}
%\addtolength{\headwidth}{\marginparsep} %these change header-rule width
%\addtolength{\headwidth}{\marginparwidth}
\lhead{PHY521/MAT597}
\chead{} 
\rhead{Mathematical Physics} 
\lfoot{} 
\cfoot{\thepage} 
\rfoot{} % !! Remember to change the problem set number
\renewcommand{\headrulewidth}{.3pt} 
%\renewcommand{\footrulewidth}{.3pt}
\setlength\voffset{0in}
%\setlength\textheight{648pt}

\begin{document}
%\input{\filepath/titlepage.tex}
%\maketitle


\title{Mathematical Physics}
\author{Lectures by Professor Michael Aizenman \\ Scribes: Holden Lee and Kiran Vodrahalli}
% INCLUDE Scribes: Holden Lee & Kiran Vodrahalli
\maketitle

%\tableofcontents

\startcontents
\printcontents{ }{-1}{}

\chapter*{Introduction}
Notes from Michael Aizenman's class ``Mathematical Physics" at Princeton in Spring 2016. 

This class is of interest to both physicists and mathematicians. Several recent Fields medals are for work related to these topics.

%don't assume knowledge
I plan to cover the following topics. The focus is on \vocab{Topics in Mathematical Statistic Mechanics}.
\begin{enumerate}
\item
\textbf{The statistical mechanic perspective}: systems can be described at the microscopic level with many degrees of freedom. We observe their collective behavior and find emergent behavior.
\item
\textbf{Thermodynamics principles}:
%fasci
%macroscopic description of physical system 
Intellectually preceding statistical mechanics is thermodynamics, a field of physics which emerged through experimental and intellectual work trying to understand what is happening with the transfer of heat. A couple of principles emerged. This framework is more appropriate to the macroscopic description of physical systems.
In departure from mechanics, which cares about equalities like $F=ma, E=mc^2$, a unique thing about thermodynamics is that its key principle is in inequality: entropy increases.
\[
\De S\ge 0
\]
%Entropy comes up as a emergent phenomenon from the statistical mechanics description of reality.
\item
\vocab{The emergence of thermodynamics from statistical mechanics via the equidistribution assumption and the ``large deviation theory"}. %vadhan - abel prize for large deviation theory.
I would like to discuss the emergence of thermodynamics from statistical mechanics. Mathematicians formalized the theory but the concepts were introduced earlier by physicists. These principles led Boltzmann to introduce these ideas and it took a while for physics to absorb the ideas.\footnote{This may have led to his premature death via suicide.}
\item 
\vocab{Phase transitions}: %Consider $H_2O$. You can control it by temperature and pressure. 
When you have a system, say $\text{H}_2\text{O}$, which you can control with temperature and pressure, you can induce changes in state. Continuous change in control parameters results in discrete jumps in the result. 

\item 
\vocab{Critical phenomena, critical exponents, universality classes}. Phase transitions are fascinating since there are interesting critical phenomena which are characterized by critical exponents, which turn out (this is one of the suprising discoveries experimentally) to result in universality classes of critical phenomena. Systems are macroscopically different, but the singularities you observe are given by the same power laws.
%Wilson.
\item
\vocab{Exact solution of the 2-D Ising model}. %From the perspective of physics. 
Mathematically, 3 is the hardest dimension to comprehend.
\begin{enumerate}
\item
1-D is solvable: correlations can be described by Markov chains, and can be computed. 
\item
In 2-D, the conformal group is very important. It gives many constraints on critical behavior, leading to a rich behavior. 
\item
In 3-D, this does not apply except for ongoing work finding consequences in 3-D from results in 2-D. 
\item
Anything with $\ge 4$ dimensions gets simpler. High dimensions are characterized by the fact that loop effects do not play a large role. In 1-D a simple random walk is recurrent. In sufficient high dimensions, a simple random walk is not recurrent.
The infinite-dimensional case reduces to models on trees; high dimensions exhibit the same behavior as infinite dimensions.
\end{enumerate}
In coding theory and discrete math people study phase transitions of different graphs and are interested in the same topics.

The 2D Ising model is usually a highly specialized topic, but we will do it in a way done a bit differently from normal. We would like to do this through ``graph zeta functions''. What I found fascinating about this topics is that there are a lot of connections to other topics. There are analogues of the Riemann zeta function on graphs, and there is a relationship. It is actually one of the simplest paths to tackle this proof by. 

\item
\vocab{Stochastic geometry behind correlation functions at criticality}: in the Ising model we have a collection of spin variables $\si_x=\pm1$ for $x\in \Z^d$. The spins are correlated in such a way that agreement among neighbors is encouraged. Thus there are correlations which spread through the system. 

It is interesting and powerful to represent this correlation between spins is via a ``shadow system'' in which you play the following game: decompose the collection of spins at random into connected clusters. You see spin values but you don't see who is connected to whom. %: But for each such decomposition the spins are connected on each cluster. 
An analogy is that students form cliques in class, and then each clique chooses what to do and votes unanimously. So if you just saw the voting pattern, you would see some cliques, but the nature of correlations among the votes become transparent if you know the clusters. 

The states for critical Ising models become larger, but also become fractal. Fractal geometric objects can be used to explain the structure. There is an interesting fractal geometry which tells us about correlation functions.

We will introduce all of this later in much detail from the ground up. This is related to percolation.

\item \vocab{Scaling limits}: Finally, if time permits, I would like to discuss scaling limits. You may want to know properties of a substance when you have something like a lattice that is very fine-grained, a statistic mechanical system which macroscopically is described by a myriad of variables. There are techniques of describing the relevant quantities from the macroscopic perspective, and this becomes relevant for describing the critical state. This provides a fascinating link between statistical mechanics and field theory. It's related to quantum field theory, except we are now in the Euclidean regime. There are some related results for quantum spin systems. 


\end{enumerate}
%random operators
%weekly packages

In a way we have a mouthful here, and one could probably give a full course on each of these topics. I will try not to be exhaustive, as we may not progress far. Previously I gave a course on random operators. Each week would cover some area of this subject, saying enough so you have a glimpse of the essence and get some comfort, and then moving on. It's now in a book format, and I would count it as a success if out of these lectures, something similar would emerge. The idea is not to be exhaustive, but to give enough key results. There is much more to be said, but I do not suppose to cover it all.

There is a broad spectrum of references, none of which I'll follow exclusively. 
\begin{itemize}
\item
Sacha Friedli and Yvan Velenik's online book project (on mathematical statistical mechanics) at \url{http://www.unige.ch/math/folks/velenik/smbook/index.html}. I recommend this for people totally new to the subject.
\item
David Ruelle formulated models and basic results mathematically (late 70's). His book helped physicists organize their thoughts. It became outdated quickly, but remains a good starting point and reference for the formalism.
\end{itemize}



%\setcounter{chapter}{-1}

\input{2-2.tex}
\input{2-4.tex}
\input{2-9.tex}
\input{2-11.tex}
\input{2-16.tex}
\input{2-18.tex}
\input{2-23.tex}
\input{2-25.tex}
\input{3-1.tex}
\input{3-3.tex}
\input{3-8.tex}
\input{3-10.tex}
\input{3-22.tex}
\input{3-24.tex}
\input{3-29.tex}
\input{3-31.tex}
\input{4-5.tex}
\input{4-7.tex}
\input{4-12.tex}
\input{4-14.tex}
\input{4-19.tex}
\input{4-21.tex}
%INSERT_HERE

%\appendix
%\input{distribution_chapters/a.tex}

%\bibliographystyle{plain}
%\bibliography{\filepath/refs}

\printnomenclature
\printindex
\end{document}